---
title: "A6_LI_ZHIJUN_Alignments"
author: "Zhijun Li"
date: '2022-03-01'
output: pdf_document
---

### GitHub Link: https://github.com/zazauwu/A6_Alignments

```{r message=FALSE, warning=FALSE}
# load the required packages
library(dplyr)
library(BiocManager)
library(Biostrings)
library(genbankr)
library(rentrez)
```

# Sequence Analysis
## Input the sequence
```{r}
#load the human isolate, unknown sequence
unknseq <- "ATGTCTGATAATGGACCCCAAAATCAGCGAAATGCACCCCGCATTACGTTTGGTGGACCCTCAGATTCAA
CTGGCAGTAACCAGAATGGAGAACGCAGTGGGGCGCGATCAAAACAACGTCGGCCCCAAGGTTTACCCAA
TAATACTGCGTCTTGGTTCACCGCTCTCACTCAACATGGCAAGGAAGACCTTAAATTCCCTCGAGGACAA
GGCGTTCCAATTAACACCAATAGCAGTCCAGATGACCAAATTGGCTACTACCGAAGAGCTACCAGACGAA
TTCGTGGTGGTGACGGTAAAATGAAAGATCTCAGTCCAAGATGGTATTTCTACTACCTAGGAACTGGGCC
AGAAGCTGGACTTCCCTATGGTGCTAACAAAGACGGCATCATATGGGTTGCAACTGAGGGAGCCTTGAAT
ACACCAAAAGATCACATTGGCACCCGCAATCCTGCTAACAATGCTGCAATCGTGCTACAACTTCCTCAAG
GAACAACATTGCCAAAAGGCTTCTACGCAGAAGGGAGCAGAGGCGGCAGTCAAGCCTCTTCTCGTTCCTC
ATCACGTAGTCGCAACAGTTCAAGAAATTCAACTCCAGGCAGCAGTAGGGGAACTTCTCCTGCTAGAATG
GCTGGCAATGGCGGTGATGCTGCTCTTGCTTTGCTGCTGCTTGACAGATTGAACCAGCTTGAGAGCAAAA
TGTCTGGTAAAGGCCAACAACAACAAGGCCAAACTGTCACTAAGAAATCTGCTGCTGAGGCTTCTAAGAA
GCCTCGGCAAAAACGTACTGCCACTAAAGCATACAATGTAACACAAGCTTTCGGCAGACGTGGTCCAGAA
CAAACCCAAGGAAATTTTGGGGACCAGGAACTAATCAGACAAGGAACTGATTACAAACATTGGCCGCAAA
TTGCACAATTTGCCCCCAGCGCTTCAGCGTTCTTCGGAATGTCGCGCATTGGCATGGAAGTCACACCTTC
GGGAACGTGGTTGACCTACACAGGTGCCATCAAATTGGATGACAAAGATCCAAATTTCAAAGATCAAGTC
ATTTTGCTGAATAAGCATATTGACGCATACAAAACATTCCCACCAACAGAGCCTAAAAAGGACAAAAAGA
AGAAGGCTGATGAAACTCAAGCCTTACCGCAGAGACAGAAGAAACAGCAAACTGTGACTCTTCTTCCTGC
TGCAGATTTGGATGATTTCTCCAAACAATTGCAACAATCCATGAGCAGTGCTGACTCAACTCAGGCCTAA"

#use regular expression 'gsub' to remove the 'carriage return' and 'newline' special character
unknseq <- gsub("[\r\n]", "", unknseq)

unknseq
```

## Generate alignments
```{r}
# pairwise alignments
library(annotate)
useqBLAST <- blastSequences(paste(unknseq),as = 'data.frame',
                            hitListSize = 20, timeout = 600)
```

```{r}
# multiple alignments
library(ape)
# create a DNAbin object
useqHitsDF <- data.frame(ID = useqBLAST$Hit_accession, # specifying an ID column
                         Seq = useqBLAST$Hsp_hseq,
                         stringsAsFactors = FALSE)
```

```{r}
# length of each sequence
useqBLAST$Hit_len
```
The 20 sequences have similar number of base pairs.

## Determine if it is human or other organism
```{r}
# read in the obtained 20 sequences from GenBank using the read.Genbank()
useqHitSeqs <- read.GenBank(useqBLAST$Hit_accession)

# take a look at the species
attr(useqHitSeqs,"species")
```

The isolated sequence is identified as Severe acute respiratory syndrome-related coronavirus 2 instead of human.

```{r}
# convert DNAbin to a DNAStringSet for an alignment
library(Biostrings)
CovDNAstring <- useqHitsDF$Seq %>% 
  as.character %>% # convert to strings
  lapply(., paste0, collapse = "") %>%  # collaspe each sequence to a single string
  unlist %>% # flatten list to a vector
  DNAStringSet # convert the vector to the required DNAStringSet object
```

```{r}
# give each sequence a unique names
names(CovDNAstring) <- paste(1:nrow(useqHitsDF),useqHitsDF$ID,sep="_")
```

```{r}
# use MUSCLE (MUltiple Sequence Comparison by Log-Expectation) to align the sequences
library(muscle)
# create a DNAMultipleAlignment object
CovAlign <- muscle::muscle(stringset = CovDNAstring, quiet = T)

CovAlign
```

```{r}
# convert the DNAMultipleAlignment object to a DNAbin
CovAlignBin <- as.DNAbin(CovAlign)

#perform a series of diagnostics on a DNA alignement
checkAlignment(CovAlignBin)
```

```{r}
SeqLen <- as.numeric(lapply(CovDNAstring, length))

# plot the distribution of sequence length
library(ggplot2)
qplot(SeqLen) + theme_bw()
```
Based on the alignment and the distribution, it looks like there is neither distinct gap nor substitution across all the 20 subject sequences, hence, there is no need to remove any sequence.

## Build a phylogeny
```{r}
# make distance matrix for tree
CDM <- dist.dna(CovAlignBin, model = "K80")
CDMmat <- as.matrix(CDM)

# rearrange CDMmat to a 'linear' matrix
library(reshape2)
PDat <- melt(CDMmat)

ggplot(data = PDat, aes(x = Var1, y = Var2, fill = value)) + 
  geom_tile() + 
  scale_fill_gradientn(colours=c("white","blue","green","red")) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))
```
Figure 1. The variation among the 20 subject sequences based on distance matrix. All of them share complete similarity with each other.

# Report
Given the results from BLAST and alignment, it should be concerned that the human isolated sequence is from the coronavirus SARS-CoV-2 which can trigger serve respiratory symptoms.
```{r}
# create a phylogenetic tree using the Neighbour-Joining (NJ) approach
CovTree <- nj(CDMmat)

# plot the phylogenetic tree
library(ggtree)
ggtree(CovTree)
```
Figure 2. Phylogenetic tree of the 20 sequences. It suggests that these sequences are closely related and fall into the same strain.

```{r}
# remove the branch length info to focus on the relationships
ggtree(CovTree, branch.length='none', layout="circular") + geom_tiplab()
```
Figure 3. Relationship among the 20 sequences. It suggests that each sequence, though of the same SARS-CoV-2 virus strain, contains a number of differences.


```{r}
# save the tree
write.tree(CovTree,"A6_LI_ZHIJUN_Cov2_tree.tre")
```




